Equalizer Programming Guide\\ {\footnotesize\sffamily\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-10-08 Developer/ProgrammingGuide/paper.tex
Programming Guide:  update to v0.9.1
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\title{Equalizer Programming Guide\\
  {\footnotesize\sffamily\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
}
\date{
  \vfill
  \vfill
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vfill
  \vfill
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9 for Equalizer v0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
using multiple processes for cluster-based execution. Equalizer provides
the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event handling
for portability. It is a `GLUT on steroids', providing parallel and
distributed execution, scalable rendering features and fully
customizable event handling.

@@

It is recommended to have a compiled version of Equalizer and the
examples for this Programming Guide. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
should be executed once to get familar with the capabilities of the
\textsf{eqPly} example.

Compiling Equalizer is as simple as running \textsf{make} on Linux or
building the Equalizer solution on Windows. On Mac OS X, some
prerequisites have to be installed before running \textsf{make}, as
explained in \textsf{README.Darwin}.

@@

The application connects to an Equalizer server, which chooses a
configuration for the application. It also provides a render client, to
be launched by the server. The application reacts on events and controls
the rendering.

@@

The render client implements --obviously-- the rendering part of an
application. Its execution is passive, completely driven by rendering
tasks received from the server. The tasks are executed by calling the
appropriate task methods (see \sref{ssTaskMethods}) in the correct
thread and context.

The application might be a rendering client, in which case it can also
contribute to the rendering. If it does not implement any render
client-related code it is reduced to be the application's 'master'
process without any OpenGL windows.

The rendering client can be the same executable as the application, as
is the case with \textsf{eqPly}. When it is started as a render client,
the Equalizer initialization routine does not return and takes over the
control by calling the render client task methods. Complex real-world
applications often implement a separate, light-weight rendering client.

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the
website\footnote{\link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe, window
and channel task methods are executed from the pipe thread for threaded
pipes or from the main application thread for non-threaded
pipes\footnote{see also
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

Compounds have a frustum description, which is inherited to the
children. The frustum can be overridden by children, but this is rarely
used since there is no use case for this behaviour. The frustrum can be
specified as a wall, which is completely defined by the bottom-left,
bottom-right and top-left coordinates relative to the origin, or as a
projection, which is defined by the position and head-pitch-roll
orientation of the projector, as well as the horizontal and vertical
field-of-view and distance to the projection wall. \fig{fFrustra}
illustrates the wall and projection frustum parameters.

@@

Compounds have output and input frames which configure the recomposition
of the resulting pixels from the source compound channels. An output
frame connects to an input frame of the same name, and transports the
selected frame buffer data from the output channel to the input
channel. This composition process is extensively described in
\sref{sCompositing}.

@@

In this section the source code of \textsf{eqPly} is discussed in
detail, and relevant design decision and remarks are raised.

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type by windowing systems. \fig{fUml} shows the
most important Equalizer classes and how they are subclassed in the
\textsf{eqPly} example.

@@

The main function starts off with parsing the command line into the
\textsf{LocalInitData} data structure, which in part will be distributed
to all render client nodes. For actual command line parsing is done by
the \textsf{LocalInitData} class and will be discussed there:

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server location. Furthermore, a node factory is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

The node factory is used by Equalizer to create the object instances for
the rendering entities. Each of the classes inherits from the same type
provided by Equalizer in the \textsf{eq} namespace. The provided
\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
object, thus making it possible to selectively subclass individual
entity types. For each rendering resource used in the configuration, one
C++ object will be created:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
is an \textsf{eqNet::Node}. The underlying network distribution in
Equalizer is a peer-to-peer network structure of
\textsf{eqNet::Node}s. The application programmer rarely is aware of the
classes in the \textsf{eqNet} namespace, but both the
\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
initialization of nodes creates a local listening socket, so that the
node, and therefore the \textsf{eq::Client} can communicate over the
network with other nodes, such as the server and the rendering clients.

@@

After it has finished, the application and Eqply is deinitialized and
the \textsf{main} function returns:

@@

In the \textbf{eqPly} case, the application is also the render client. It has
three run-time behaviours:

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later implementations to
provide additional hints and information to the server for choosing a
configuration. The configuration is created using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making q cast necessary:

@@

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measures and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} does return after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all init task methods were successful. The \textsf{EQLOG} macro allows
topic-specific logging. The numeric topic values are specified in the
respective \textsf{log.h} header files:

@@

When the configuration was successfully initialized, the actual main
loop is executed. The main loop runs until the user exits the
configuration or a maximum number of frames, specified as a command-line
argument, has been rendered. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::finishFrame}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames for which the
tasks are already queued in the node and pipe threads for
execution. This enables overlapped execution and minimizes idle
times. The first \textsf{latency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical)
rendering tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

@@

In the second and third case, when the executable is used as a render
client, \textsf{Client::initLocal} never returns. Therefore the
application's main loop is never executed. In order to keep the client
resident, the \textsf{eqPly} example overrides the client loop to keep
it running beyond one configuration run:

@@

Distributed object are created by subclassing from
\textsf{eqNet::Object}. Distributed objects can be static (immutable) or
dynamic. Dynamic objects are versioned.

@@

\subsubsection{InitData - a Static Distributed Object}

@@

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData} or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be outlayed
in one contiguous block of memory. The second approach is used
otherwise.

@@

The data output and input streams perform no type checking on the
variables. It is the application's responsibility to match the order and
types of variables during serialization and deserialization exactly.

@@

The frame data is layed out in one anonymous structure in
memory. It also does not track changes since it is relatively small in
size and changes frequently. Therefore, for the instance and delta
date are the same and set in the constructor:

@@

The configuration is driving the applications rendering, that is, it is
responsible for updating the data based on received events, requesting
new frames to be rendered and to provide the render clients with the
necessary data.

@@

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This is necessary to allow a speedy startup of
the configuration on large-scale graphics clusters. On the other hand,
it means that initialization functions are called even if the parent's
initialization has failed.

@@

The frame data is registered first, since its identifier is transmitted
using the initialization data, which is registered afterwards. The
identifier of the initialization data is transmitted to the render
client nodes using the \textsf{initID} parameter of
\textsf{eq::Config::init}. Equalizer will pass this identifier to all
\textsf{configInit} calls of the respective objects:

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering, and has been solved already by a number
of implementations\footnote{VRCO Trackd, VRPN, etc.}. The example
code in \textsf{eqPly} is one reference implementation for the
integration of such a tracking library:

@@

The rendering frames are issued by the application. The \textsf{Config}
only overrides \textsf{startFrame} in order to update the its data
before forwarding the start frame request to the \textsf{eq::Config}.

@@

The camera position is updated and the frame data is commited, which
generates a new version. This version is passed to the rendering
callbacks and will be used to synchronize the frame data to the state
belonging to the current frame:

@@

{\footnotesize\begin{lstlisting}
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->type )
    {
        case eq::ConfigEvent::WINDOW_CLOSE:
            _running = false;
            return true;

        [...]

        case eq::ConfigEvent::POINTER_MOTION:
            if( event->pointerMotion.buttons == eq::PTR_BUTTON_NONE )
                return true;

@@

During node initialization the initialization data is mapped to
a local instance using the passed identifier from
\textsf{Config::init}. The model is loaded based on the filename in the
initialization data. No pipe, window or channel tasks methods are
executed before \textsf{Node::configInit} has returned.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the execution within the thread.

Likewise, \textsf{Config::finishFrame} causes the invokation of the
\textsf{frameFinish} task methods. These task methods unlock their
parents by calling \textsf{finishFrame}.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. The window and channel synchronization is similar and
omitted for simplicity. The \textsf{eqPly} example does not override
\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
vital for the execution that \textsf{Node::startFrame} or
\textsf{Node::finishFrame} are called, respectively. The default
implementation of the node task methods does take care of that.

@@

The \textsf{eqPly} examples allows selecting the window system using a
command line option. Therefore the implementation of
\textsf{selectWindowSystem} is overwritten, and return the specified
window system, if it is valid:

@@

Parts of the Carbon API used for window and event handling in the AGL
window system are not thread safe. The application has to call
\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call,
and \textsf{eq::Global::leaveCarbon} afterwards. These functions should
be used only during window initialization and exit, not during
rendering\footnote{for various reasons \textsf{enterCarbon} might block
  up to 50 milliseconds}. Carbon call in the window event handling
routine \textsf{Window::processEvent} are thread-safe. Please contact
the Equalizer developer mailing list if you need to use Carbon calls on
a per-frame basis.

@@

The Equalizer window holds an OpenGL drawable and rendering
context. Subclassed windows should maintain all data specific to the
OpenGL context. The Equalizer window creation routines share the OpenGL
context with the first window of the pipe, this allowing the reuse of
OpenGL objects, e.g., display lists and textures.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the pipe's window system. The default implementations
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

@@

The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail below. It is used in conjunction
with a reference pointer here, since it is potentially `owned' by
multiple windows at the same time.

The logo texture is loaded from the file system and bound to a texture
ID used later by the channel for rendering. The details of this code are
omitted here, since they are pretty straight-forward and not
Equalizer-specific.

The window exit deallocates all OpenGL objects if the state object is
about to be disposed. The object manager does not delete the object in
its destructor, since it does not know if an OpenGL context is still
current. In any case, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

@@

\subsubsection{Object Manager}

@@

The state object in \textsf{eqPly} gathers all rendering state, which
includes an object manager for object state handling. This state object
can easily be replaced to use the rendering code in a stand-alone
application, e.g., for an GLUT-based renderer.

@@

The OM is a template class. The template type is the key type, by which
objects can be identified. The same key is used by all contexts to get
the OpenGL name of an object. In \textsf{eqPly}, a key of type
\textsf{const void *} is used. The rendering code uses the address of
the data item to be rendered to obtain the associated OpenGL object.

@@

Currently handling for display lists, textures and buffers is
implemented. For each object, the following functions are available:

>>>>

\title{Equalizer Programming Guide\\
  {\footnotesize\sffamily\mdseries
    \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf}
    {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}
}
\author{Stefan Eilemann\\{\footnotesize Eyescale Software GmbH}
}
\date{
  \vspace{4cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{DRAFT}\\[\medskipamount]
  Version 0.9.1 for Equalizer v0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\tableofcontents
\thispagestyle{empty}
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.1   & Oct 8, 2007  & review pass\\
    0.9     & Oct 7, 2007  & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer implements the minimum necessary layer to build parallel and
scalable OpenGL applications. The programmer structures the application
so that the OpenGL rendering can be executed in parallel, potentially
using multiple processes for cluster-based execution. Equalizer provides
the domain-specific parallel rendering know-how and abstracts
configuration, threading, synchronization, windowing and event
handling. It is a `GLUT on steroids', providing parallel and distributed
execution, scalable rendering features and fully customizable event
handling.

@@

It is recommended to have a compiled version of Equalizer and the
examples when working with this Programming Guide. The Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
should be executed once to get familar with the capabilities of the
\textsf{eqPly} example.

Compiling Equalizer is as simple as running \textsf{make} on Linux or
building the Equalizer Visual Studio 2005 solution on Windows. On Mac OS
X 10.4 (Tiger), some prerequisites have to be installed before running
\textsf{make}, as explained in \textsf{README.Darwin}. Mac OS X 10.5
Leopard does have all the prerequisites installed by default.

@@

The application connects to an Equalizer server, which chooses a
configuration for the application. Furthermore, the application also
provides its render client, which will be controlled by the server. The
application reacts on events, updates its database and controls the
rendering.

@@

The render client implements --obviously-- the rendering part of an
application. Its execution is passive, that is, it has no main loop and
is completely driven by rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context.

The application might be a rendering client, in which case it can also
contribute to the rendering. If it does not implement any render
client-related code it is reduced to be the application's `master'
process without any OpenGL windows.

The rendering client can be the same executable as the application, as
it is the case with \textsf{eqPly}. When it is started as a render client,
the Equalizer initialization routine does not return and takes over the
control by calling the render client task methods. Complex real-world
applications often implement a separate, light-weight rendering client.

@@

The application subclasses Equalizer objects and overrides virtual
functions to implement certain functionality, e.g., the application's
OpenGL rendering in \textsf{eq::Chan\-nel::frameDraw}. These task
methods are in concept similar to C function callbacks. The
\textsf{eqPly} section will discuss the most important task methods. A
full list can be found on the website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/taskMethods.html}}.

\if 0
\subsection{Thread Safety}

Using threading correctly in OpenGL-based applications is easy with
Equalizer. Equalizer creates one rendering thread for each graphics
card. All task methods for a pipe, and therefore all OpenGL commands,
are executed from this thread. This threading model is the OpenGL
`threading model', which maintains one current context for each
thread. It is difficult, if not impossible, to render from different
threads to the same context.

The rendering threads concurrently render the application's data
base. The data base should be accessed in a read-only fashion during
rendering, as it is the case with all modern scene graphs.

The rendering threads run asynchronously to the application's main
thread. Depending on the configuration's latency, the fall $n$ frames
behind the last frame finished by the application thread.

TBD support and explain per-node frame sync.
\fi

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU). In
the current implementation it is also one operating system thread,
unless the pipe's thread hint is set to \textsf{false}. All pipe, window
and channel task methods are executed from the pipe thread pipes or from
the main application thread for non-threaded pipes\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

Compounds have a frustum description, which is inherited to the
children. The frustum can be overridden by children, but this is rarely
used since there is no use case for this behaviour.

The frustrum can be specified as a wall, which is completely defined by
the bottom-left, bottom-right and top-left coordinates relative to the
origin, or as a projection, which is defined by the position and
head-pitch-roll orientation of the projector, as well as the horizontal
and vertical field-of-view and distance of the projection
wall. \fig{fFrustra} illustrates the wall and projection frustum
parameters.

@@

Compounds have output and input frames which configure the recomposition
of the resulting pixels from the source compound channels. An output
frame connects to an input frame of the same name, which transports the
selected frame buffer data from the output channel to the input
channel. This composition process is extensively described in
\sref{sCompositing}.

@@

In this section the source code of \textsf{eqPly} is discussed in
detail, and relevant design decision, caveat and remarks are raised.

All classes in the example are in the \textsf{eqPly} namespace to avoid
type name ambiguities, in particular for the \textsf{Window} class which
is frequently used as a type in the global namespace by windowing
systems. \fig{fUml} shows the most important Equalizer classes and how
they are subclassed by the \textsf{eqPly} example.

@@

The main function starts off with parsing the command line into the
\textsf{LocalInitData} data structure, part of which will be distributed
to all render client nodes. The actual command line parsing is done by
the \textsf{LocalInitData} class and will be discussed in
\sref{sInitData}:

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

The node factory is used by Equalizer to create the object instances for
the rendering entities. Each of the classes inherits from the same type
provided by Equalizer in the \textsf{eq} namespace. The provided
\textsf{eq::NodeFactory} base class instantiates a 'plain' Equalizer
object, thus making it possible to selectively subclass individual
entity types, as it is done by \textsf{eqHello}. For each rendering
resource used in the configuration, one C++ object will be created:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is an \textsf{eq::Client}, which
in turn is an \textsf{eqNet::Node}. The underlying network distribution
in Equalizer is a peer-to-peer network structure of
\textsf{eqNet::Node}s. The application programmer rarely is aware of the
classes in the \textsf{eqNet} namespace, but both the
\textsf{eq::Client} and the server are \textsf{eqNet::Node}s. The local
initialization of nodes creates a local listening socket, so that the
node, and therefore the \textsf{eq::Client} can communicate over the
network with other nodes, such as the server and the rendering clients.

@@

After it has finished, the application and \textsf{eqPly} is
de-initialized and the \textsf{main} function returns:

@@

In the case of \textsf{eqPly} the application is also the render
client. The \textsf{eqPly} executable has three run-time behaviours:

\begin{enumerate}
  \item \textbf{Application}: The executable started by the user, which
    is the controlling entity in the rendering session.
  \item \textbf{Auto-launched render client}: The typical render client,
    started by the server. The server starts the executable with special
    parameters, which cause \textsf{Client::initLocal} to never
    return. During exit, the server terminates the process. By default,
    the server starts render client using \textsf{ssh}.
  \item \textbf{Resident render client}: Manually pre-started render
    client, listening on a specified port for server commands. This mode
    is selected using the command-line option \textsf{--eq-client} and
    potentially \textsf{--eq-listen} and \textsf{-r}\footnote{see \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

The second step is to ask the server for a configuration. The
\textsf{ConfigParams} are a placeholder for later Equalizer
implementations to provide additional hints and information to the
server for choosing a configuration. The configuration chosen by the
server is created locally using
\textsf{NodeFactory::createConfig}. Therefore it is of type
\textsf{eqPly::Config}, but the return value is \textsf{eq::Config},
making cast necessary:

@@

Finally it is time to initialize the configuration. For statistics, the
time for this operation is measures and printed. During initialization
the server launches and connects all render client nodes, and calls the
appropriate initialization task methods, as explained in later
sections. \textsf{Config::init} does return after all nodes, pipes,
windows and channels are initialized. It returns \textsf{true} only if
all initialization task methods were successful.

The \textsf{EQLOG} macro allows topic-specific logging. The numeric
topic values are specified in the respective \textsf{log.h} header
files, and logging for various topics is enables using the environment
variable \textsf{EQ\_LOG\_TOPICS}:

@@

When the configuration was successfully initialized, the main rendering
loop is executed. The main loop runs until either the user exits the
configuration or a maximum number of frames has been rendered, specified
as a command-line argument. The latter is useful for benchmarks. The
\textsf{Clock} is reused for measuring the overall performance. A new
frame is started using \textsf{Config::startFrame} and a frame is
finished using \textsf{Config::finishFrame}.

@@

\begin{wrapfigure}{r}{.6\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapped execution in the node and pipe threads and minimizes idle
times. The first \textsf{nLatency Config::finishFrame} return
immediately, since they have no frame to synchronize
upon. \fig{fSyncAsync} shows the execution of (hypothetical) rendering
tasks without latency~(\fig{fSync}) and with a latency of
one~(\fig{fAsync}).

With \textsf{eqPly} a speedup of 15\% has been observed on a five-node
rendering
cluster when using a latency of one instead of no latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}.

@@

In the second and third usage case of the \textsf{eqPly}, when the
executable is used as a render client, \textsf{Client::initLocal} never
returns. Therefore the application's main loop is never executed. In
order to keep the client resident, the \textsf{eqPly} example overrides
the client loop to keep it running beyond one configuration run:

@@

Distributed objects are created by subclassing from
\textsf{eqNet::Object}. Distributed objects can be static (immutable) or
dynamic. Dynamic objects are versioned.

@@

\subsubsection{\label{sInitData}InitData - a Static Distributed Object}

@@

On the application side, the class \textsf{LocalInitData} subclasses
\textsf{InitData} to provide the command line parsing and to set the
initial values. The render nodes only instantiate the distributed part
in \textsf{InitData}

A static distributed object either has to provide a pointer and size to
its data using \textsf{setInstanceData}, or it has to implement
\textsf{getInstanceData} and \textsf{applyInstanceData}. The first
approach can be used if all distributed member variables can be outlayed
in one contiguous block of memory. The second approach is used
otherwise.

@@

The data output and input streams perform no type checking on the data
stream. It is the application's responsibility to match the order and
types of variables during serialization and deserialization exactly.

@@

The \textsf{eqPly} frame data is layed out in one anonymous structure in
memory. It also does not track changes since it is relatively small in
size and changes frequently. Therefore, the instance and delta data are
the same and set in the constructor:

@@

The configuration is driving the application's rendering, that is, it is
responsible for updating the data based on received events, requesting
new frames to be rendered and to provide the render clients with the
necessary data.

@@

The tasks are executed by the node and pipe threads in parallel. The
parent's initialization methods are always executed before any child
initialization method. This parallelization is necessary to allow a
speedy startup of the configuration on large-scale graphics clusters. On
the other hand, it means that initialization functions are called even
if the parent's initialization has failed.

@@

The frame data is registered first, since its identifier is transmitted
using the initialization data. The initialization data is registered
afterwards so that it can transmit the frame data's identifier. The
identifier of the initialization data is transmitted to the render
client nodes using the \textsf{initID} parameter of
\textsf{eq::Config::init}. Equalizer will pass this identifier to all
\textsf{configInit} calls of the respective objects:

@@

If the configuration was initialized correctly, the configuration tries
to set up a tracking device for head tracking. Equalizer does not
provide extensive support for tracking device, as this is an orthogonal
problem to parallel rendering. Tracking device support has been solved
already by a number of implementations\footnote{VRCO Trackd, VRPN,
  etc.}, which can easily be integrated with Equalizer. The example code
in \textsf{eqPly} is one reference implementation for the integration of
such a tracking library:

@@

The rendering frames are issued by the application. The \textsf{Config}
only overrides \textsf{startFrame} in order to update its data before
forwarding the start frame request to the \textsf{eq::Config}.

@@

The camera position is updated and the frame data is commited, which
generates a new version for this object. This version is passed to the
rendering callbacks and will be used by the rendering threads to
synchronize the frame data to the state belonging to the current frame:

@@

{\footnotesize\begin{lstlisting}
bool Config::handleEvent( const eq::ConfigEvent* event )
{
    switch( event->type )
    {
        [...]
        case eq::ConfigEvent::POINTER_MOTION:
            if( event->pointerMotion.buttons == eq::PTR_BUTTON_NONE )
                return true;

@@

During node initialization the init data is mapped to a local instance
using the passed identifier from \textsf{Config::init}. The model is
loaded based on the filename in the initialization data. No pipe, window
or channel tasks methods are executed before \textsf{Node::configInit}
has returned.

@@

The application has extended control over the task synchronization
during a frame. Upon \textsf{Config::startFrame}, Equalizer invokes the
\textsf{frameStart} task methods of the various entities. The entity
unlock all its children by calling \textsf{startFrame}, e.g.,
\textsf{Node::frameStart} has to call \textsf{Node::startFrame} in order
to unlock the pipe threads. Note that certain \textsf{startFrame} calls,
e.g., \textsf{Window::startFrame}, are currently empty since the
synchronization is implicit due to the sequential execution within the
thread.

Likewise, \textsf{Config::finishFrame} causes the invokation of the
\textsf{frameFinish} task methods. These task methods unlock their
parents by calling \textsf{releaseFrame}.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. The window and channel synchronization is similar and
omitted for simplicity. The \textsf{eqPly} example does not override
\textsf{Node::frameStart} or \textsf{frameFinish}, but it is absolutely
vital for the execution that \textsf{Node::startFrame} or
\textsf{Node::releaseFrame} are called, respectively. The default
implementation of the node task methods do take care of that.

@@

The \textsf{eqPly} examples allows selecting the window system using a
command line option. Therefore the implementation of
\textsf{selectWindowSystem} is overwritten, and returns the specified
window system, if it is supported:

@@

Parts of the Carbon API used for window and event handling in the AGL
window system are not thread safe. The application has to call
\textsf{eq::Global::enterCarbon} before any thread-unsafe Carbon call,
and \textsf{eq::Global::leaveCarbon} afterwards. These functions should
be used only during window initialization and exit, not during
rendering. For various reasons \textsf{enterCarbon} might block up to 50
milliseconds. Carbon call in the window event handling routine
\textsf{Window::processEvent} are thread-safe, since the global carbon
lock is set in this method. Please contact the Equalizer developer
mailing list if you need to use Carbon calls on a per-frame basis.

@@

The Equalizer window holds an OpenGL drawable and rendering
context. Subclassed windows should maintain all data specific to the
OpenGL context. The Equalizer window creation routines shares the OpenGL
context with the first window of the pipe, thus allowing the reuse of
OpenGL objects between all windows of one pipe, e.g., display lists and
textures.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the pipe's window system. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which is emant to perform the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail in \sref{sObjectManager}. It is
used in conjunction with a reference pointer here, since it is
potentially `owned' by multiple windows at the same time.

The logo texture is loaded from the file system and bound to a texture
ID used later by the channel for rendering. The details of this code are
omitted here, since they are pretty straight-forward OpenGL code and not
Equalizer-specific.

The window exit deallocates all OpenGL objects when the state object is
about to be disposed. The object manager does not delete the object in
its destructor, since it does not know if an OpenGL context is still
current. In any case, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

@@

\subsubsection{\label{sObjectManager}Object Manager}

@@

The state object in \textsf{eqPly} gathers all rendering state, which
includes an object manager for object state handling. This state object
can easily be replaced to use the rendering code in a stand-alone
application, e.g., for a GLUT-based renderer.

@@

The OM is a template class. The template type is the key type, by which
objects are identified. The same key is used by all contexts to get the
OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
  void *} is used. The rendering code uses the address of the data item
to be rendered as the key to obtain the associated OpenGL object.

@@

Currently handling for display lists, VBO's, textures and shaders is
implemented. For each object, the following functions are available:
